---
title: "Heart Failure Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Heart Failure Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data\>*

### Reference:

*\<Larxel. (2019). Heart Failure Prediction Dataset. Retrieved from Kaggle
https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r Load dataset}
# Load dataset
heart_failure_data <- read.csv("heart_failure.csv", colClasses = c(
  age = "numeric",
  anaemia = "factor",
  creatinine_phosphokinase = "numeric",
  diabetes = "factor",
  ejection_fraction = "numeric",
  high_blood_pressure = "factor",
  platelets = "numeric",
  serum_creatinine = "numeric",
  serum_sodium = "numeric",
  sex = "factor",
  smoking = "factor",
  time = "numeric",
  DEATH_EVENT = "factor"
))

# Display the structure of the dataset
str(heart_failure_data)

# View the first few rows of the dataset
head(heart_failure_data)

# View the dataset in a separate viewer window
View(heart_failure_data)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Count of unique values in each factor variable
frequency <- sapply(heart_failure_data[, sapply(heart_failure_data, is.factor)], table)
print("Measures of Frequency:")
print(frequency)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
print("\nMeasures of Central Tendency:")
# Mean
means <- colMeans(heart_failure_data[, sapply(heart_failure_data, is.numeric)], na.rm = TRUE)
print("Mean:")
print(means)
# Median
medians <- apply(heart_failure_data[, sapply(heart_failure_data, is.numeric)], 2, median, na.rm = TRUE)
print("Median:")
print(medians)
# Mode (for factor variables)
modes <- sapply(heart_failure_data[, sapply(heart_failure_data, is.factor)], function(x) names(which.max(table(x))))
print("Mode:")
print(modes)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
print("\nMeasures of Distribution:")
# Standard Deviation
sds <- apply(heart_failure_data[, sapply(heart_failure_data, is.numeric)], 2, sd, na.rm = TRUE)
print("Standard Deviation:")
print(sds)
# Range
ranges <- apply(heart_failure_data[, sapply(heart_failure_data, is.numeric)], 2, function(x) diff(range(x, na.rm = TRUE)))
print("Range:")
print(ranges)
# Interquartile Range (IQR)
iqrs <- apply(heart_failure_data[, sapply(heart_failure_data, is.numeric)], 2, IQR, na.rm = TRUE)
print("Interquartile Range (IQR):")
print(iqrs)
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
print("\nMeasures of Relationship:")
# Correlation Matrix for numeric variables
correlation_matrix <- cor(heart_failure_data[, sapply(heart_failure_data, is.numeric)], use = "complete.obs")
print("Correlation Matrix:")
print(correlation_matrix)
```

## ANOVA
```{r ANOVA}
# Perform ANOVA for numeric variables across different factors
anova_results <- lapply(heart_failure_data[, sapply(heart_failure_data, is.numeric)], function(var) {
  anova_result <- lapply(heart_failure_data[, sapply(heart_failure_data, is.factor)], function(factor_var) {
    anova_result <- anova(lm(var ~ factor_var, data = heart_failure_data))
    return(anova_result)
  })
  return(anova_result)
})

# Print ANOVA results
print("ANOVA Results:")
print(anova_results)
```

## Plots
```{r Plots}
# Load required libraries
library(ggplot2)

# Univariate Plots
# Histogram of age
ggplot(heart_failure_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency")

# Bar plot of anaemia
ggplot(heart_failure_data, aes(x = anaemia)) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Frequency of Anaemia", x = "Anaemia", y = "Frequency") +
  scale_x_discrete(labels = c("No", "Yes"))

# Multivariate Plot
# Scatter plot of age vs. serum_creatinine colored by DEATH_EVENT
ggplot(heart_failure_data, aes(x = age, y = serum_creatinine, color = DEATH_EVENT)) +
  geom_point() +
  labs(title = "Age vs. Serum Creatinine by Death Event", x = "Age", y = "Serum Creatinine", color = "Death Event") +
  scale_color_manual(values = c("blue", "red")) +
  theme(legend.title = element_blank())

# Box plot of ejection_fraction by diabetes
ggplot(heart_failure_data, aes(x = diabetes, y = ejection_fraction, fill = diabetes)) +
  geom_boxplot() +
  labs(title = "Ejection Fraction by Diabetes Status", x = "Diabetes", y = "Ejection Fraction", fill = "Diabetes")

# Pairwise scatter plot of numeric variables
pairs(~ age + creatinine_phosphokinase + ejection_fraction + platelets + serum_creatinine + serum_sodium, data = heart_failure_data)

```

# Preprocessing and Data Transformation
## Missing Values
```{r missing values}
# Check for missing values
missing_values <- colSums(is.na(heart_failure_data))
print("Presence of Missing Values:")
print(missing_values)
```

# Training Model
## Data Splitting
```{r Data Splitting}
## Load required package
library(caret)

# Data Splitting
set.seed(123)  # For reproducibility
train_index <- createDataPartition(heart_failure_data$DEATH_EVENT, p = 0.7, list = FALSE)
train_data <- heart_failure_data[train_index, ]
test_data <- heart_failure_data[-train_index, ]

# Print the dimensions of the training and testing sets
cat("Training set dimensions:", dim(train_data), "\n")
cat("Testing set dimensions:", dim(test_data), "\n")
```

## Bootstrapping
```{r Bootstrapping}
# Bootstrapping
boot_data <- createDataPartition(1:nrow(heart_failure_data), p = 0.8, list = FALSE)
bootstrapped_samples <- heart_failure_data[boot_data, ]
```

## Cross-validation
```{r Cross-validation}
# Cross-validation
# Basic cross-validation
cv_model <- train(DEATH_EVENT ~ ., data = heart_failure_data, method = "glm", trControl = trainControl(method = "cv"))
print("Basic Cross-validation:")
print(cv_model)
```

## Training Different Models
```{r Training Different Models}
# Load required libraries
library(caret)
library(randomForest)
library(glmnet)

# Set seed for reproducibility
set.seed(123)

# Define training control
train_control <- trainControl(method = "cv", number = 5)

# Train Random Forest model
rf_model <- train(DEATH_EVENT ~ ., data = heart_failure_data, method = "rf", trControl = train_control)
print("Random Forest Model:")
print(rf_model)

# Train glmnet model
glmnet_model <- train(DEATH_EVENT ~ ., data = heart_failure_data, method = "glmnet", trControl = train_control)
print("GLMNET Model:")
print(glmnet_model)

# Train Generalized Linear Model (GLM)
glm_model <- train(DEATH_EVENT ~ ., data = heart_failure_data, method = "glm", trControl = train_control)
print("GLM Model:")
print(glm_model)
```

## Compare performance
```{r Compare Performance}
# Compare model performance using resamples
resamps <- resamples(list(Random_Forest = rf_model, GLMNET = glmnet_model, GLM = glm_model))
summary(resamps)
```

## Saving Model
```{r Saving Model}
# Load the saved Random Forest model
loaded_heart_failure_rf_model <- readRDS("./models/heart_failure_rf_model.rds")

# Sample new data 
new_heart_failure_data <- data.frame(
  age = c(55, 65, 75),  
  anaemia = factor(c(0, 1, 0)),  
  creatinine_phosphokinase = c(7861, 146, 582),  
  diabetes = factor(c(0, 0, 0)),  
  ejection_fraction = c(38, 20, 20),  
  high_blood_pressure = factor(c(0, 0, 1)),  
  platelets = c(263358.03, 162000, 265000),  
  serum_creatinine = c(1.1, 1.3, 1.9),  
  serum_sodium = c(136, 129, 130),  
  sex = factor(c(1, 1, 1)),  
  smoking = factor(c(0, 1, 0)),  
  time = c(6, 7, 4),  
  DEATH_EVENT = factor(c(1, 1, 1))
)

# Use the loaded model to make predictions for new data
predictions <- predict(loaded_heart_failure_rf_model, new_heart_failure_data)

# Print predictions
print(predictions)

```

