---
title: "Heart Failure Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Heart Failure Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data\>*

### Reference:

*\<Larxel. (2019). Heart Failure Prediction Dataset. Retrieved from Kaggle
https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r Load dataset}
# Load dataset
heart_failure_data <- read.csv("heart_failure.csv", colClasses = c(
  age = "numeric",
  anaemia = "factor",
  creatinine_phosphokinase = "numeric",
  diabetes = "factor",
  ejection_fraction = "numeric",
  high_blood_pressure = "factor",
  platelets = "numeric",
  serum_creatinine = "numeric",
  serum_sodium = "numeric",
  sex = "factor",
  smoking = "factor",
  time = "numeric",
  DEATH_EVENT = "factor"
))

# Display the structure of the dataset
str(heart_failure_data)

# View the first few rows of the dataset
head(heart_failure_data)

# View the dataset in a separate viewer window
View(heart_failure_data)
```

